---
output: html_document
---
# Synopsis

The purpose of this project is to demonstrate ability to collect, work with, and clean 
a data set. The goal is to prepare tidy data that can be used for later analysis. 
Required to submit: 1) a tidy data set as described below, 2) a link to a Github 
repository with script for performing the analysis, and 3) a code book that describes 
the variables, the data, and any transformations or work that performed to clean up the data called CodeBook.md.


# Data

Companies like Fitbit, Nike, and Jawbone Up are racing to develop the most advanced algorithms to attract new users. The data linked to from the course website represent data collected from the accelerometers from the Samsung Galaxy S smartphone. A full description is available at the site where the data was obtained:

http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones

Here are the data for the project:

https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip

# Assignment

Student have to create one R script called run_analysis.R that does the following:

* Merges the training and the test sets to create one data set.
* Extracts only the measurements on the mean and standard deviation for each measurement.
* Uses descriptive activity names to name the activities in the data set.
* Appropriately labels the data set with descriptive variable names.
* From the data set in step 4, creates a second, independent tidy data set with 
the average of each variable for each activity and each subject.

# GitHUB link to the project files

* [R script, Code book, and tidy data file](https://github.com/rblenin/Getting-and-Cleaning-Data-Course-Project.git)

# 1. Downloading and unzipping dataset
```{r eval=FALSE}
if(!file.exists("./data")) {
  dir.create("./data")
}
fileUrl <- "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
download.file(fileUrl, destfile = "./data/Dataset.zip", method = "curl")
unzip(zipfile = "./data/Dataset.zip", exdir = "./data")
```

# 2. Merging the training and test sets to create one data set
## 2.1 Reading training files

```{r eval=FALSE}
xTrain <- read.table("./data/UCI HAR Dataset/train/X_train.txt")
yTrain <- read.table("./data/UCI HAR Dataset/train/y_train.txt")
subjectTrain <- read.table("./data/UCI HAR Dataset/train/subject_train.txt")
```

## 2.2 Reading testing files
```{r eval=FALSE}
xTest <- read.table("./data/UCI HAR Dataset/test/X_test.txt")
yTest <- read.table("./data/UCI HAR Dataset/test/y_test.txt")
subjectTest <- read.table("./data/UCI HAR Dataset/test/subject_test.txt")
```

## 2.3 Reading feature file
```{r eval=FALSE}
features <- read.table("./data/UCI HAR Dataset/features.txt")
```

## 2.4 Reading activity label file
```{r eval=FALSE}
activityLabels <- read.table("./data/UCI HAR Dataset/activity_labels.txt")
```

## 2.5 Assigning column names
```{r eval=FALSE}
colnames(xTrain) <- features[,2]
colnames(yTrain) <- "activityId"
colnames(subjectTrain) <- "subjectId"

colnames(xTest) <- features[,2]
colnames(yTest) <- "activityId"

colnames(activityLabels) <- c("activityId", "activityType")
```

## 2.6 Merging training and test data sets
```{r eval=FALSE}
train <- cbind(yTrain, subjectTrain, xTrain)
test <- cbind(yTest, subjectTest, xTest)
myData <- rbind(train, test)
```

# 3. Extracting only the measurements on the mean and standard deviation for each measurement.
## 3.1 Extracting variable names that contain the words 'mean' and 'std'
```{r eval=FALSE}
subFeatures<-features$V2[grep("mean\\(\\)|std\\(\\)", features$V2)]
extractFeatures <- c("activityId", "subjectId", as.character(subFeatures))
```

## 3.2 Extracting variables identified in Step 3.1
```{r eval=FALSE}
myData <- subset(myData, select = extractFeatures)
````

# 4. Using descriptive activity names to name the activities in the data set
## 4.1 Adding a column 'activityType' which names the activity according to the 'activityId'
```{r eval=FALSE}
myData <- merge(myData, activityLabels, by = "activityId", all.x = TRUE)
```

## 4.2 Removing the 'activityId' column and moving the last column 'activityType' to the first column
```{r eval=FALSE}
myData <- myData[,2:ncol(myData)]
colId <- grep("activityType", names(myData))
myData <- myData[, c(colId, (1:ncol(myData))[-colId])]
```

# 5. Appropriately labeling the data set with descriptive variable names
```{r eval=FALSE}
names(myData)<-gsub("^t", "time", names(myData))
names(myData)<-gsub("^f", "frequency", names(myData))
names(myData)<-gsub("Acc", "Accelerometer", names(myData))
names(myData)<-gsub("Gyro", "Gyroscope", names(myData))
names(myData)<-gsub("Mag", "Magnitude", names(myData))
names(myData)<-gsub("BodyBody", "Body", names(myData))
```

# 6. From the data set in the previous step, creating a second, independent tidy data set with the average of each variable for each activity and each subject.
## 6.1 Finding average of each variable for each acitivityType and each subject 
```{r eval=FALSE}
myData2 <- aggregate(. ~activityType + subjectId, myData, mean)
```

## 6.2 Interchanging 'subjectId' and 'activityType' columns so that 'subjectId' and 'activityType' columns are the first and second columns in the data set, and ordering the data set based on 'subjectId' and 'activityType'
```{r eval=FALSE}
colId <- grep("subjectId", names(myData2))
myData2 <- myData2[, c(colId, (1:ncol(myData2))[-colId])]
myData2 <- myData2[order(myData2$subjectId, myData2$activityType), ]
```

## 6.3 Writing the tidy data set to a text file 'tidyData.txt'
```{r eval=FALSE}
write.table(myData2, file = "tidyData.txt", row.name = FALSE)
```
